Cognitive Analysis

### 2024-07-25 Manuscript Analysis/manuscript_analysis.Rmd
WM pupil analysis

library(tidyverse)
library(rio)
library(glmmTMB)
library(janitor)
library(ggpubr)
library(patchwork)
library(gt)
library(gtsummary)
library(ggridges)
library(performance)
library(sjPlot)
library(ggcorrplot)

@@

# Listener Demographics

The listeners used in this study are the same from the parent study. In brief, the demographic information from these listeners are below.

@@

# Listener Cognition

@@

>>>>

library(tidyverse)
library(rio)
library(glmmTMB)
library(janitor)
library(ggpubr)
library(patchwork)
library(gt)
library(gtsummary)
library(ggridges)
library(performance)
library(sjPlot)
library(ggcorrplot)
library(mgcv) # install.packages("mgcv")
library(itsadug) # install.packages("itsadug")

@@

# Listener Characteristics

## Demographic Information

The listeners used in this study are the same from the parent study. In brief, the demographic information from these listeners are below. A total of 34 young adult listeners participated in the experiment.

@@

## Listener Cognition

The information below reports the mean age-corrected standardized scores on the subtests from the NIH Toolbox cognition battery (working memory, inhibitory control, cognitive flexibility, and processing speed).

@@

# Color Palette for Figures
my_pal <- c("#033F63", "#2CB5AE", "#28666E", "#4BAA9B", "#49E9B9", "#57A773")

cog |>
  dplyr::select(subject, ends_with("_c")) |>
  dplyr::rename("Working Memory" = list_sort_c,
                "Inhibitory Control" = flanker_c,
                "Cognitive Flexibility" = card_sort_c,
                "Processing Speed" = pattern_c) |>
  tidyr::pivot_longer(cols = 'Working Memory':'Processing Speed',
                      names_to = "cog_measure",
                      values_to = "score") |>
  ggplot() +
  aes(x = score,
      y = cog_measure,
      color = cog_measure,
      fill = cog_measure) +
  geom_density_ridges(jittered_points = T,
                      position = position_points_jitter(width = 0.05, height = 0),
                      point_shape = '|', point_size = 3, point_alpha = 1, alpha = 0.5) +
  labs(x = "Score", y = "", color = "Cognitive Measure", fill = "Cognitive Measure") +
  scale_color_manual(values = c(my_pal[3],my_pal[4], my_pal[5], my_pal[6])) +
  scale_fill_manual(values = c(my_pal[3],my_pal[4], my_pal[5], my_pal[6])) +
  theme_classic() +
  theme(legend.position = "none")

# Removing unneeded items

rm(cog_table, demo_table, demo)

# Pupil Dilation

# Selecting needed variables from phrase repetition accuracy df

acc <- phrase_acc |>
  dplyr::select(subject, trial, target_number, correct_words_initial, rep_acc)

## Merging acc df with pupil dilation df

pupil <- pupil |>
  # Merging phrase accuracy information with df
  dplyr::left_join(acc, by = c("subject", "trial")) |>
  # Filter out trials with missing phrase repetition data |>
  dplyr::filter(!is.na(rep_acc)) |>
  # Merging cognitive data
  dplyr::left_join(cog, by = "subject")

## Base Model

Final Model from parent study

pupil <- pupil |>
  dplyr::mutate(speaker = factor(speaker, levels = c("Control", "ALS")),
                rep_acc = factor(rep_acc, levels = c("accurate", "inaccurate")),
                phrase = as.factor(code),
                listener = as.factor(subject),
                condition = case_when(speaker == "Control" & rep_acc == "accurate" ~ "Control Accurate",
                                      speaker == "Control" & rep_acc == "inaccurate" ~ "Control Inaccurate",
                                      speaker == "ALS" & rep_acc == "accurate" ~ "ALS Accurate",
                                      TRUE ~ "ALS Inaccurate"),
                condition = factor(condition, 
                                   levels = c("Control Accurate", "Control Inaccurate", "ALS Accurate", "ALS Inaccurate"))) |>
  dplyr::group_by(listener, speaker) |>
  dplyr::mutate(num_points = n()) |>
  dplyr::mutate(start_event = c(TRUE, rep(FALSE, each = (num_points - 1)))) |>
  ungroup()

m0_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(m0_pupil)

m0_prediction <- plot_smooth(m0_pupil,
            view = "time_norm",
            plot_all = "condition",
            rug = F, se = 1, rm.ranef = T)

m0_predicted_values <- m0_prediction[[1]][c("condition", "time_norm", "fit", "ul", "ll")] |>
  dplyr::mutate(rep_acc = ifelse(grepl("inaccurate", condition, ignore.case = T), "inaccurate", "accurate"),
                speaker = ifelse(grepl("Control", condition, ignore.case = T), "Control", "ALS"),
                speaker = factor(speaker, levels = c("Control", "ALS")))

m0_predicted_values |>
  ggplot() +
  aes(x = time_norm,
      y = fit,
      color = speaker,
      fill = speaker) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 1910, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 2910, ymin = 0, ymax = 200, alpha = 0.15) +
  coord_cartesian(ylim = c(-10, 250), xlim = c(0, 4910)) +
  facet_wrap("rep_acc") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Normalized Time (ms)",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Speaker",
       fill = "Speaker",
       linetype = "Speaker") +
  scale_color_manual(values = c(my_pal[1], my_pal[5]), labels = c("Neurotypical", "ALS")) +
  scale_fill_manual(values = c(my_pal[1], my_pal[5]), labels = c("Neurotypical", "ALS"))

## Working Memory

### Initial Model

wm_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  te(time_norm, list_sort_c, k = 20) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(wm_pupil)

#### Model Check

gam.check(wm_pupil)

acf_resid(wm_pupil)

#### Model Comparison

compareML(m0_pupil, wm_pupil, signif.stars = T, suggest.report = T)

### Model 1

wm_pupil_1 <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  te(time_norm, list_sort_c, by = speaker, k = 20) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(wm_pupil_1)

#### Model Check

gam.check(wm_pupil_1)

acf_resid(wm_pupil_1)

#### Model Comparison

compareML(m0_pupil, wm_pupil_1, signif.stars = T, suggest.report = T)

```
### Model 2

wm_pupil_2 <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  te(time_norm, list_sort_c, by = condition, k = 20) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(wm_pupil_2)

#### Model Check

gam.check(wm_pupil_2)

acf_resid(wm_pupil_2)

```
#### Model Comparison

compareML(m0_pupil, wm_pupil_2, signif.stars = T, suggest.report = T)
